OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2022-01-18 paper/response-to-reviewers.md

**Figure 1: W-Z are designated as the query sequences in the caption. Z
does not appear in the distance threshold table at the top of the figure
and is not assigned to an OTU at the end of Cycle 4. From Line 115
(“Alternatively, a sequence will remained unassigned…”) the reader can
infer that sequence Z is different enough from the others that it was
designated a singleton (outside of the distance threshold discussed in
lines 111-113). If I’m correct, I encourage the authors to make this
point explicitly in Figure 1’s caption.**

That’s correct, we added a sentence to the caption as you suggested.

**Figure 3: I suggest changing the label for “de novo” to “OptiClust (de
novo)” or similar to emphasize that OptiClust was being used for those
benchmark tests.**

In Figure 3, we show results from *de novo* clustering with both the
OptiClust algorithm in mothur (red triangles) and VSEARCH (blue
triangles). Changing the label to “OptiClust (*de novo*)” would not be
accurate, as VSEARCH does not use the OptiClust algorithm. Instead, we
changed the red color label from “mothur” to “OptiClust (*de novo*) or
OptiFit” in Figure 3.

### 2022-01-18 paper/paper.Rmd
Fix typo: OptiFit uses more RAM than OptiClust

figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))

figs <- function(fig_name, attribute, include_figures = params$include_figures) {
  value <- figures_meta[[fig_name]][[attribute]]
  if (attribute == 'cap') {
    value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
  } else if (attribute == 'dim' & !isTRUE(include_figures)) {
    value <- c(0.1,0.1)
  } else {
    value <- eval(parse(text = value))
  }
  return(value)
}

render_figure <- function(fig_name, include_figures = params$include_figures) {
  if (isTRUE(include_figures)) {
    figs(fig_name, 'path') %>%
      tiff::readTIFF() %>%
      grid::grid.raster()
  } else {
    ggplot() + theme_void()
  }
}

@@

To test how OptiFit performs for reference-based clustering, we clustered each
dataset to three databases of reference OTUs: 
the Greengenes database v13_8_99 [@desantis_greengenes_2006], 
the SILVA non-redundant database v132 [@quast_silva_2013], 
and the Ribosomal Database Project (RDP) v16 [@cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% using the V4 region of each
sequence (see Figure \ref{fig:workflow}).
After trimming to the V4 region, the databases contained 
`r dn_dbs %>% pull('nseqs')` unique sequences and produced _de novo_ MCC scores 
of `r dn_dbs %>% pull('med_mcc')` for `r dn_dbs %>% pull('refname')`, 
respectively.
Clustering query sequences with OptiFit to Greengenes and SILVA in closed
reference mode performed similarly, with median MCC scores of 
`r closed_fit_gg_mcc` and `r closed_fit_silva_mcc` respectively, while the median
MCC was `r closed_fit_rdp_mcc` when clustering to RDP (Figure
\ref{fig:results_sum}; "db: Greengenes", "db: SILVA", and "db: RDP").
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc` (Figure \ref{fig:results_sum}; "_de novo_").
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes,
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering with OptiClust, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
render_figure('workflow')
```

@@

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH in terms of MCC score, but VSEARCH was able to
cluster `r frac_vsearch_vs_fit`% more query sequences than OptiFit to the
Greengenes reference database (Figure \ref{fig:results_sum}).
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of reference and query sequences in an OTU.
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively.
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and
`r sec_vsearch_vs_open_fit_db`% on average, respectively.
Thus, the more stringent OTU definition employed by OptiFit, which prefers the
query sequence to be similar to all other sequences in the OTU rather than to only one
sequence, resulted in fewer sequences being clustered to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split)
(Figure \ref{fig:results_sum}; "self-split").
OTU quality was similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences clustered to the Greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and
`r sec_open_fit_split_vs_clust`%, respectively.
Random access memory (RAM) usage was similar, with OptiFit requiring slightly 
more RAM in gigabytes than OptiFit. 
Open and closed reference OptiFit required `r mem_open_fit_split_vs_clust`% and
`r mem_closed_fit_split_vs_clust`% more RAM than OptiClust, respectively 
(data not shown).
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
render_figure('results_split')
```

@@

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To identify the best reference size, reference sets with 10% to 90% of the
sequences were created, with the remaining sequences used for the query
(Figure \ref{fig:results_split}).
OTU quality was remarkably consistent across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation (i.e. the standard deviation divided by the mean) of
`r cv_fit_split_mcc_human_simple` for the MCC score across all fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` seconds with
10% of sequences in the reference and `r sec_fit_split_human_simple_9` seconds
with 90% of sequences in the reference.
The RAM usage was virtually the same across reference fraction sizes, with a
coefficient of variation of `r cv_fit_split_mem_human_simple` for the human
dataset (data not shown).
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference.
These trends held for the other datasets as well.
Thus, the reference fraction did not affect OTU quality in terms of MCC score
nor the memory usage, but did affect the run time and the fraction of sequences
that mapped during the closed reference clustering.

>>>>

figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))
include_figures <- params$include_figures
figs <- function(fig_name, attribute, incl_figs = include_figures) {
  value <- figures_meta[[fig_name]][[attribute]]
  if (attribute == 'cap' | attribute == 'caption') {
    value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
  } else if (attribute == 'dim' & !isTRUE(incl_figs)) {
    value <- c(0.1,0.1)
  } else {
    value <- eval(parse(text = value))
  }
  return(value)
}

render_figure <- function(fig_name) {
    figs(fig_name, 'path') %>%
      tiff::readTIFF() %>%
      grid::grid.raster()
}

@@

To test how OptiFit performs for reference-based clustering, we clustered each
dataset to three databases of reference OTUs: 
the Greengenes database v13_8_99 [@desantis_greengenes_2006], 
the SILVA non-redundant database v132 [@quast_silva_2013], 
and the Ribosomal Database Project (RDP) v16 [@cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% using the V4 region of each
sequence (see Figure 2).
After trimming to the V4 region, the databases contained 
`r dn_dbs %>% pull('nseqs')` unique sequences and produced _de novo_ MCC scores 
of `r dn_dbs %>% pull('med_mcc')` for `r dn_dbs %>% pull('refname')`, 
respectively.
Clustering query sequences with OptiFit to Greengenes and SILVA in closed
reference mode performed similarly, with median MCC scores of 
`r closed_fit_gg_mcc` and `r closed_fit_silva_mcc` respectively, while the median
MCC was `r closed_fit_rdp_mcc` when clustering to RDP 
(Figure 3; "db: Greengenes", "db: SILVA", and "db: RDP").
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc` (Figure 3; "_de novo_").
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes,
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering with OptiClust, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

@@

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH in terms of MCC score, but VSEARCH was able to
cluster `r frac_vsearch_vs_fit`% more query sequences than OptiFit to the
Greengenes reference database (Figure 3).
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of reference and query sequences in an OTU.
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively.
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and
`r sec_vsearch_vs_open_fit_db`% on average, respectively.
Thus, the more stringent OTU definition employed by OptiFit, which prefers the
query sequence to be similar to all other sequences in the OTU rather than to only one
sequence, resulted in fewer sequences being clustered to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split) (Figure 3; "self-split").
OTU quality was similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences clustered to the Greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and
`r sec_open_fit_split_vs_clust`%, respectively.
Random access memory (RAM) usage was similar, with OptiFit requiring slightly 
more RAM in gigabytes than OptiClust. 
Open and closed reference OptiFit required `r mem_open_fit_split_vs_clust`% and
`r mem_closed_fit_split_vs_clust`% more RAM than OptiClust, respectively 
(data not shown).
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

@@

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To identify the best reference size, reference sets with 10% to 90% of the
sequences were created, with the remaining sequences used for the query
(Figure 4).
OTU quality was remarkably consistent across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation (i.e. the standard deviation divided by the mean) of
`r cv_fit_split_mcc_human_simple` for the MCC score across all fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` seconds with
10% of sequences in the reference and `r sec_fit_split_human_simple_9` seconds
with 90% of sequences in the reference.
The RAM usage was virtually the same across reference fraction sizes, with a
coefficient of variation of `r cv_fit_split_mem_human_simple` for the human
dataset (data not shown).
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference.
These trends held for the other datasets as well.
Thus, the reference fraction did not affect OTU quality in terms of MCC score
nor the memory usage, but did affect the run time and the fraction of sequences
that mapped during the closed reference clustering.
